---
output:
  pdf_document: default
  html_document: default
---
```library(tidyverse)```
---
title: "Lab 4"
author:  Nandini Parmar 305561701
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float: yes
---
\fontsize{10}{11}

&nbsp;&nbsp;

```{r datetime, echo=FALSE, eval=TRUE}
library(tidyverse)

cat("Date last run:", format(Sys.time(), "%Y-%m-%d"))
cat("\n")
cat("Hello World!")
cat("\n")


options(width=90)

```


&nbsp;&nbsp;


\newpage


# Your Work

Make sure to edit the "author" information in the YAML header near the top to include your name and UID.

Complete/answer the following.


1 --- Does our NFL team data represent “stacked” data? Why or why not?

The NFL team data does not represent stacked data because the rows are independent from each other and the data does not form a stack that is dependent on what is before it.

2 --- Each team played 17 regular season games.  Calculate the average points scored per game for each team.  Create a scatterplot showing avg team points versus avg passer ratings with the least squares line ...

> 2a --- Does this plot visually look similar to its counterpart using season total points above? ...

> 2b --- Is the Pearson's correlation coefficient the same or different?  Why?



```{r, echo=TRUE, message=FALSE, eval=TRUE, results='show', fig.width=10,fig.height=5}
library(xtable)
library(tidyverse)
library(readxl)
library(ggthemes)

xdf <- read_excel("/Users/apurvashah/Documents/GitHub/stats10/lab4/NFL_offense_passing_2021.xlsx", sheet=1, col_names = TRUE, skip = 1)
head(xdf, n=6)

x <- data.frame(Team = xdf$Rate, avg = xdf$totalPoints)
head(x)
x$avg = round(x$avg/17, 2)

# head(x)

xmeanPR <- mean(x$Team)
xmeanPR

xmeanPnts <- mean(x$avg)
xmeanPnts


##### sample standard deviation of rating
xsdPR <- sd(x$Team)
xsdPR

##### sample standard deviation of total points
xsdPnts <- sd(x$avg)
xsdPnts

n <- nrow(xdf)
n

###### sample covariance
xcov <- sum((x$Team - xmeanPR) * (x$avg - xmeanPnts) ) / (n - 1)
xcov


###### sample correlation
xcorr <- xcov / (xsdPR * xsdPnts)
xcorr

###### sample regression line slope
xb1 <- xcorr * xsdPnts / xsdPR
xb1

###### sample regression line y-intercept
xb0 <- xmeanPnts - xb1 * xmeanPR
xb0

graph <- ggplot(data = x,  aes(Team, avg)) + geom_point(size = 1, color = "blue") + 
  labs(title = "NFL 2021: Average Points Scored / Game vs. QB Passer Rating",
       x = "Team Passer Rating", y = "Average Points Scored / Game") + theme_stata() +
  geom_abline(slope=xb1, intercept=xb0, color = "maroon")

graph



```

The graph look very similar to the one above because the only difference would be in the how all of the data is scaled. 
The correlation coefficient is the same because the data is the same just scaled differently. 

&nbsp;



3 --- Read in the NHL team-game data, and examine the relationship between assists and goals (for the regression part, goals are the outcome, assists are the predictor, i.e., regress goals on assists).



```{r, echo=TRUE, message=FALSE, eval=TRUE, results='show'}

xdf <- read_tsv("/Users/apurvashah/Documents/GitHub/stats10/lab4/NHL_20202021_teamGame.tsv", col_names = TRUE)
head(xdf)

x <- data.frame(Team = xdf$goals, avg = xdf$assists)
head(x)

xmeanPR <- mean(x$Team)
xmeanPR

xmeanPnts <- mean(x$avg)
xmeanPnts


##### sample standard deviation of rating
xsdPR <- sd(x$Team)
xsdPR

##### sample standard deviation of total points
xsdPnts <- sd(x$avg)
xsdPnts

n <- nrow(xdf)
n

###### sample covariance
xcov <- sum((x$Team - xmeanPR) * (x$avg - xmeanPnts) ) / (n - 1)
xcov


###### sample correlation
xcorr <- xcov / (xsdPR * xsdPnts)
xcorr

###### sample regression line slope
xb1 <- xcorr * xsdPnts / xsdPR
xb1

###### sample regression line y-intercept
xb0 <- xmeanPnts - xb1 * xmeanPR
xb0


```

&nbsp;


4 --- With NHL team-game data, can we argue that increasing assists causes more goals?

Although when looking at the data, we can see that when there are more assists, there generally more goals, since we did not conduct and experiment and did not use random assignment we cannot assume causality here. We can say that there is a correlation. 



&nbsp;


5 --- Back to the NFL data, can we say that improving passer rating causes more points to be scored?  

Note that 4 and 5 are intended to test your reasoning.  These questions may not be as simple as they appear.

Although we might see a trend that teams with better passer ratings cause more points to be scored we cannot say that this causes more points to be scored because we did not conduct a true experiment to test this. We cannot assume causality and at most can say that there might be a correlation between these two.





